### 2021-05-03 paper/sex_selection_spacing-final.tex
Edit for final submission

\title{Birth Spacing and Fertility in the Presence of Son Preference and Sex-Selective Abortions:
India's Experience Over Four Decades%
\protect\thanks{%
I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
I owe thanks to Monica Das Gupta, Shelly Lundberg, Daniel Rees, David Ribar, 
Hendrik Wolff, three anonymous reviewers, and seminar participants at University of 
Copenhagen, University of Michigan, University of Washington, University of Aarhus, the 
Fourth Annual Conference on Population, Reproductive Health, 
and Economic Development, and the Economic Demography Workshop for helpful 
suggestions and comments.
I would also like to thank Nalina Varanasi for research assistance.
Support for development of the method from the University of Washington Royalty 
Research Fund and the Development Research Group of the World Bank is gratefully 
acknowledged.
The views and findings expressed here are those of the author and
should not be attributed to the World Bank or any of its member countries.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
}
}

@@

\date{December 2020}

@@

% Demography abstract
\noindent 
% In India, the most substantial lengthening in Hindu women's birth intervals over the past 
% four decades has occurred for the women most likely to use sex-selective abortions. 
% For well-educated women with two girls, the median birth interval before the third child 
% has increased by nearly 15 months---and the 75th percentile interval by 21 months---as the 
% sex ratio has approached 80\% boys. 
% Because of the rapid lengthening in spacing, standard fertility rates have significantly 
% overestimated how fast cohort fertility fell. 
% Despite convergence, cohort fertility is still 10\%--20\% higher than the 
% fertility rate and above replacement level for all but the best-educated urban women. 
% With the lengthening from sex selection, some women show a reversal of the traditional 
% spacing pattern; instead of the shortest birth interval in the absence of sons, they now 
% have the longest. 
% Finally, infant mortality has declined over time for all groups, but the decline was 
% fastest for the less educated. 
% Short birth spacing is still associated with higher mortality, although considerably less 
% so for the best-educated women. 
% There is no evidence that the increasing use of sex selection is associated with higher 
% infant mortality. 
% Neither does it appear that the use of sex selection is declining.

% \noindent Over the past four decades, the Hindu women in India most likely to use 
% sex-selective abortions---well-educated women with no sons---had the most substantial 
% lengthening of birth intervals and the most biased sex ratios. 
% As a result, we now see cases that reverse the traditional spacing pattern, with some
% women with no sons having longer birth intervals than those with sons. 
% Those least likely to use sex-selective abortions---less-educated women in rural 
% areas---still follow the traditional pattern of short spacing when they have girls, with 
% only limited evidence of sex selection. 
% Because of the rapid lengthening in spacing, the standard fertility rates substantially 
% overestimated how fast cohort fertility fell. 
% Despite a convergence, cohort fertility is still 10\%--20\% higher than the fertility rate 
% and above replacement level for all but the best-educated urban women. 
% Infant mortality has declined substantially over time for all groups, with the fastest 
% decline among the less educated. 
% Short birth spacing is still associated with higher mortality, although considerably less 
% so for the best-educated women. 
% There is no evidence that repeated sex-selective abortions are associated with higher 
% infant mortality for the child eventually born. 
% Finally, it does not appear that the use of sex selection is declining.

% \noindent India has experienced an increasingly male-biased sex ratio at birth following
% the introduction of prenatal sex determination techniques. 
% I examine three questions.
% First, how did birth spacing change with this and other secular developments? 
% Second, did changes in birth spacing make us overestimate the decline in cohort fertility? 
% Finally, did the advent of sex-selective abortions change the relationship between birth 
% spacing and infant mortality? 
% Well-educated Hindu women with no sons had the most substantial lengthening of birth 
% intervals and the most male-biased sex ratios, likely from sex-selective abortions. 
% Consequently, some women with no sons now have longer birth intervals than those with sons, 
% reversing India's traditional spacing pattern.
% Less-educated women continued short spacing when they have girls, with only limited
% evidence of male-biased sex ratios. 
% Because of the overall lengthening in spacing, period fertility rates substantially 
% overestimated how fast cohort fertility fell. 
% Furthermore, predicted cohort fertility remains above replacement for all but the 
% best-educated urban women. 
% If lengthening of birth intervals occurs from repeated abortions, the associated
% short pregnancy spacing may counteract any positive effects of longer birth spacing. 
% There is no evidence of this effect on infant mortality. 
% Finally, judging from sex ratios, sex-selective abortion use is not declining.

@@

However, India also witnessed the advent of sex selection---the selective abortion of 
female fetuses based on prenatal sex determination---with the introduction of ultrasound 
in the mid-1980s.
Combined with a strong continued preference for sons, the result was a dramatic increase 
in the male--female ratio at birth
\citep{das_gupta97,Arnold2002,retherford03b,Guilmoto2012,Portner2015b,Jayachandran2017}.

% Questions addressed and why valuable
The main question I address here is how birth spacing responded to these changes, 
especially the spread of sex selection. 
The motivation is twofold.
First, past research has failed to appreciate that the use of sex-selective abortions could 
substantially increase birth spacing. 
As detailed below, it takes 6 months or more after an abortion to reach the same point in 
the next pregnancy. 
Second, other societal changes, such as women's greater educational attainment, higher 
household income, and a low and declining female labor force participation, all 
likely influence birth spacing. 
The combination of the apparent increasing use of sex selection and the other societal 
changes can significantly impact birth spacing, but we know little about how much.

@@

First, birth intervals lengthened over the four decades, with the lengthening longer, the
higher the parity, the more educated the woman, and the higher the percentile.
Well-educated women with no sons had the most substantial lengthening of birth intervals 
and the most male-biased sex ratios, both likely arising from sex-selective abortions. 
Consequently, some women with no sons now have \emph{longer} birth intervals than those 
with sons, reversing India's traditional spacing pattern. 
Less-educated women continued to have short spacing when they have girls, with only limited 
evidence of male-biased sex ratios.
The likelihood of a very short birth interval changed little across most groups.

@@

This section provides a conceptual framework for understanding changes in birth spacing.
I first introduce three potential explanations that link fertility and birth 
spacing decisions: economic conditions, investment in children, and son preference 
\citep{Casterline2016,Portner2018}.
Next, I discuss why female education, area of residence, and sex composition are the
principal factors in the empirical analyses and tie them into the three explanations.

@@

An increased desire to invest in children may also lengthen birth spacing. 
The clearest example of a positive effect of longer spacing is for health as very short 
spacing---approximately 24 months or less---leads to worse child health and mortality 
outcomes \citep{Whitworth2002,Conde-Agudelo2012}.
Whether longer spacing also improves human capital outcomes is more speculative with
mixed evidence for developed countries and no evidence for developing countries
\citep{Zajonc1976,Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.

@@

% Factors/predictions
The closest available proxies in the data for capturing the explanations linking
fertility and birth spacing are the sex composition of previous children, mothers'
education, and area of residence. 
The remainder of the section discusses how the three variables tie into the
explanations and the predicted effects.

@@

For birth interval lengths, I estimate a set of percentile durations. 
I first calculate for each woman when there is a given percentage chance that she will have 
given birth, conditional on the probability of giving birth in the spell. 
For example, with an 80\% parity progression probability, the median birth interval 
is the predicted number of months before a woman passes the 60\% mark on her survival 
curve. 
I then add nine months to account for the start of the spell. 
The reported statistic is the average of a given percentile interval across all women in 
a given sample using the individual progression probabilities as weights.

@@

These results are in line with prior research showing that falling fertility is associated 
with a lengthening of particularly the longer birth intervals, although why is still an 
unresolved question \citep{Casterline2016}.
The most educated women who already have a son is an exception to this trend; the 
probability of a third birth declined rapidly, but the birth interval lengths changed little. 
These women both have better access to modern contraceptives and are better at using 
traditional contraceptive methods \citep{Rosenzweig1989}. 
Hence, it is possible that the lengthening of birth intervals with falling fertility in 
the prior research arises because unintended births become a relatively larger proportion 
of the total number of births for a given parity.

@@

Even more striking is that most of the change took place right at the introduction of sex 
selection.
The 75th percentile birth interval with two girls increased from 48 months to 
64 months in a decade, while the other sex compositions showed a slight decrease from 
around 55 months to 54 months.
These birth spacing changes may even understate the impact because this particular group
appears to have had access to sex selection even before it became widespread, as shown by
the unequal sex ratio for the 1972--1984 period for women with two girls.

>>>>

\title{Birth Spacing and Fertility in the Presence of Son Preference and Sex-Selective Abortions:
India's Experience Over Four Decades%
\protect\thanks{%
I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
I owe thanks to Monica Das Gupta, Shelly Lundberg, Daniel Rees, David Ribar, 
Hendrik Wolff, three anonymous reviewers, and seminar participants at University of 
Copenhagen, University of Michigan, University of Washington, University of Aarhus, the 
Fourth Annual Conference on Population, Reproductive Health, 
and Economic Development, and the Economic Demography Workshop for helpful 
suggestions and comments.
I would also like to thank Nalina Varanasi for research assistance.
Support for development of the method from the University of Washington Royalty 
Research Fund and the Development Research Group of the World Bank is gratefully 
acknowledged.
The views and findings expressed here are those of the author and
should not be attributed to the World Bank or any of its member countries.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
Text and code for this paper is available at 
\href{https://github.com/cportner/sexSelectionSpacing}{https://github.com/cportner/sexSelectionSpacing}
}
}

@@

\date{May 2021}

@@

@@

However, India also witnessed the advent of sex selection---the selective abortion of 
female fetuses based on prenatal sex determination---with the introduction of ultrasound 
in the mid-1980s.
Combined with a strong, continued preference for sons, the result was a dramatic increase 
in the male--female ratio at birth
\citep{das_gupta97,Arnold2002,retherford03b,Guilmoto2012,Portner2015b,Jayachandran2017}.

% Questions addressed and why valuable
The main question I address here is how birth spacing responded to these changes, 
especially the spread of sex selection. 
The motivation is twofold.
First, past research has failed to appreciate that the use of sex-selective abortions could 
substantially increase birth spacing. 
As detailed below, after an abortion it takes 6 months or more to reach the same point in 
the next pregnancy. 
Second, other major societal changes, such as women's greater educational attainment, higher 
household income, and a low and declining female labor force participation, all 
likely influence birth spacing. 
The combination of the apparent increasing use of sex selection and societal 
changes can significantly impact birth spacing, but we know little about how much.

@@

First, birth intervals lengthened over the four decades, with the lengthening longer, the
higher the parity, the more educated the woman, and the higher the percentile of the
birth interval duration.
Well-educated women with no sons had the most substantial lengthening of birth intervals 
and the most male-biased sex ratios, both likely arising from sex-selective abortions. 
Consequently, some women with no sons now have \emph{longer} birth intervals than those 
with sons, reversing India's traditional spacing pattern. 
Less-educated women continued to have short spacing when they have girls, with only limited 
evidence of male-biased sex ratios.
The likelihood of a very short birth interval changed little across most groups.

@@

This section provides a conceptual framework for understanding changes in birth spacing.
I first introduce three potential explanations that link fertility and birth 
spacing decisions: economic conditions, investment in children, and son preference 
\citep{Casterline2016,Portner2018}.
Next, I discuss why female education, area of residence, and sex composition are the
principal factors in the empirical analyses and tie them to the three explanations.

@@

An increased desire to invest in children may also lengthen birth spacing. 
The clearest example of a positive effect of longer spacing is for health, as very short 
spacing---approximately 24 months or less---leads to worse child health and mortality 
outcomes \citep{Whitworth2002,Conde-Agudelo2012}.
Whether longer spacing also improves human capital outcomes is more speculative with
mixed evidence for developed countries and no evidence for developing countries
\citep{Zajonc1976,Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.

@@

% Factors/predictions
The closest available proxies in the data for capturing the explanations linking
fertility and birth spacing are the sex composition of previous children, mothers'
education, and area of residence. 
The remainder of the section discusses how the three variables fit into the
explanations and the predicted effects.

@@

For birth interval lengths, I estimate a set of percentile durations. 
I first calculate for each woman when there is a given percentage chance that she will have 
given birth, conditional on the probability of giving birth in the spell. 
For example, with an 80\% parity progression probability, the median birth interval 
is the predicted number of months before a woman passes the 60\% $(100-\frac{80}{2})$ 
mark on her survival curve. 
I then add nine months to account for the start of the spell. 
The reported statistic is the average of a given percentile interval across all women in 
a given sample using the individual progression probabilities as weights.

@@

These results are in line with prior research showing that falling fertility is associated 
with a lengthening of particularly the longer birth intervals, although why is still an 
unresolved question \citep{Casterline2016}.
The most educated women who already have a son is an exception to this trend; the 
probability of a third birth declined rapidly, but the birth interval lengths changed little. 
These women both have better access to modern contraceptives and are better at using 
traditional contraceptive methods \citep{Rosenzweig1989}. 
Hence, it is possible that the lengthening of birth intervals with falling fertility in 
the prior research arises because unintended births become a relatively larger proportion 
of the total number of births for a given parity and those can occur at any time.

@@

Even more striking, most of the change took place right at the introduction of sex 
selection.
The 75th percentile birth interval with two girls increased from 48 months to 
64 months in a decade, while the other sex compositions showed a slight decrease from 
around 55 months to 54 months.
These birth spacing changes may even understate the impact because this particular group
appears to have had access to sex selection even before it became widespread, as shown by
the unequal sex ratio for the 1972--1984 period for women with two girls.
